Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-04-19 WholePlotTrajectories.tex
Figures captions

To counter the remaining taxonomic uncertainty plots taxonomic
composition and diversity were analysed at the genus level, \emph{i.e.}
referring to the genus of observed or trialed botanical names. The
analysis of taxonomic and functional composition of plots along time was
visualized in a two-dimensional ordination space based on non-metric
dimensional scaling of succesive floristic or functional inventories.
Plots trajectories along time was reported comparatively to the
inventories in 1989, 5 years after disturbance, which corresponded to
first inventory with a sufficient degree of uncertainty (\textless{}30\%
of undetermined trees). The inventories dissimilarity compared to the
reference 1989 inventory was reported using occurrence-based (Jaccard)
and abundance-based (Bray-Curtis) similarity measures. The trajectory of
inventories along time was visualized with the euclidean distance in the
two-dimensional ordination space to the 1989 inventory. The functional
trajectories of the leaf and stem and life traits were also visualized
with the community weighted means (CWM), representing the average trait
value in a community weighted by relative abundance of the species
carrying each value \citep{Diaz2007, Garnier2004}. Species seed mass
correspond to 5 classes of increasing mass, the seed mass trajectories
were reported by the proportion of each class recorded in the
inventories. To determine whether communities' assembly deviated from
the expectation of a random process we compared the observed similarity
patterns to those generated by 50 repetitions of stochastic null models.
The null model for taxonomic composition randomly shuffled individuals
among plots while preserving the overall species abundance and plots'
tree density. The null model for functional composition produced random
traits assemblages by permuting the set of functional traits values
among species.

@@

Over time, 828388 trees and 591 botanical species pertaining to 223
genus and 64 botanical families were recorded. Trajectories of taxonomic
and functional composition after disturbance were examined through the
ordination of successive inventories from 1989 (5 years after
disturbance) in the two-dimensional space from NMDS analysis based on
flora inventories and the corresponding functional composition.
Classifications were performed using either abundance-based Bray-Curtis
(Figure \ref{fig:Fig1}) or incidence-based Jaccard dissimilarity (data
not shown).

Both taxonomic and functional composition were substantially affected by
disturbance, and the maximum dissimilarity to 1984 reference inventory
composition was positively correlated with the disturbance intensity.
Until 30 years after disturbance, all disturbed plots taxonomic
composition remained significantly dissimilar to that of the 1989
reference inventory composition (Figure \ref{fig:Fig1}). All plots,
though, displayed a unimodal trajectory with a the return toward the
taxonomic and functional composition of the 1989 reference inventory
suggesting a shift towards a cyclic regime (Figure \ref{fig:Fig2}). For
taxonomic composition, the maximum dissimilarity to 1984 inventory
composition was reach around 20 years after disturbance for the T2 and
T3 plots and for one T1 plot while control and other T1 plots continued
to increase (Figure \ref{fig:Fig2}.a). For functional composition, the
maximum dissimilarity to 1984 inventory composition was reach from 15 to
20 years after disturbance for the T2 and T3 plots while T1 and control
plots continued to increase (Figure \ref{fig:Fig2}.b). The functional
composition distancing from the reference inventory was also positively
correlated with disturbance intensity and similarly stabilized or
reduced (for 5 of T1, T2 and T3 plots) from 20 years after disturbance
(Figure \ref{fig:Fig2}).

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/Fig1-1}

@@

\caption{Trajectories of the plots in terms of **(a)** flora composition and **(b)** functional composition regarding the 6 leaf and stem functional traits,the maximum allometric height and seed mass class in the two-dimensional space from the NMDS performed for the 30 years after disturbance. Distance matrix for NMDS were computed from the Bray-curtis dissimilarity between successive inventories. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3).}\label{fig:Fig1}
\end{figure*}

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/Fig2-1}

@@

\caption{Trajectories of the distance to initial condition of the 30 sampled years in the two-dimensional space from the NMDS of **(a)** taxonomic composition at genus level and **(b)** functional composition. Distance are abundance-based Bary-Curtis metric. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3). The 0.025 and 0.975 percentile correspond to the variance observed for 50 iteration of the taxonomic uncertainty propagation and functional trait filling processes. }\label{fig:Fig2}
\end{figure}

@@

For all plots the trajectories of the community weighted means (CWM)
were drawn for the 8 functional and life history traits (Leaf thickness,
chlorophyll content, toughness and specific area,wood specific gravity
and barck thickness and seed mass and maximum adult height) (Figure
\ref{fig:Fig3}).

@@

Except for leaf cholorphyll content, which displayed higher difference
between plots than among treatments which may be due to the completeness
of dataset, all CWM trajectories corresponded to significant changes
after disturbance. All functional traits and seed mass proportions
displayed a unimodal trajectories but with different times at maximum
and different values 30 years after disturbance. The weighted means of
communities specific maximum height at adult stage (\emph{Hmax}), leaf
toughness (\emph{L\_toughness}) and wood specific gravity (\emph{WD})
remained significantly lower than their initial value and than these of
the control plots (Figure \ref{fig:Fig3}). The weighted means of bark
thickness (\emph{Bark\_thick}) similarly remained substantilly higher
than initially for all disturbed plots while the specfic leaf area
(\emph{SLA}) had almost recovered its initial value and this of the
undisturbed plots at the end of the experiment.

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/Fig3-1}

@@

\caption{Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, *L_thickness*, chlorophyll content, *L_chloro*, toughness, *L_toughness* and specific area, *SLA*), 2 stem traits (wood specific gravity, *WD*, and bark thickness, *Bark-thick*) and one life trait (Specific maximum height at adult stage,*Hmax*). Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.}\label{fig:Fig3}
\end{figure*}

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/Fig4-1}

\caption{Trajectories of seed mass classes proportions over 30 years after disturbance, corresponding to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. No gap filling process was applied in this case. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:Fig4}
\end{figure*}

@@

Trajectories of Richness, Shannon and Simpson taxonomic diversity were
examined at genus level in relation to the 1989 inventories (5 years
after disturbance) (Figure \ref{fig:Fig5}).

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/Fig5-1}

@@

\caption{Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities **(a)** Richness, **(b)** Shannon and **(c)** Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:Fig5}
\end{figure*}

While for undisturbed plots the Richness, Shannon and Simpson diversity
remained comparable to the values in 1989, the trajectories of all
disturbed plots showed significant increase positively correlated with
the disturbance intensity (\(\rho_{spearman}^{Shannon}=0.87\), and
\(\rho_{spearman}^{Simpson}=0.88\)) (Figure \ref{fig:Fig5}). The
Richness was less affected by the initial disturbance except for the
higher intensity (T3 plots) (Figure \ref{fig:Fig5}). The Shannon and
Simpson diversities however first significantly increased, illustrating
an increasing evenness, before reaching a maximum around 15 years after
disturbance for all plots except one from T1 treatment and one from T2
treatment (Figure \ref{fig:Fig5}).

@@

For all undisturbed plots the Rao diversity remained comparable to the
values in 1989, the trajectories of all disturbed plots showed
significant increase positively correlated with the disturbance
intensity (Figure \ref{fig:Fig6}). Although a first increase of plots
functional diversity that was all the more important than initial
disturbance was intense, 30 after disturbance the diversity was similar
among disturbed plots and significantly higher than this of control
plots.

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/Fig6-1}

@@

\caption{Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes.}\label{fig:Fig6}
\end{figure}

@@

The monitoring 30 years of disturbed forest communities confirmed the
limited direct impact of low disturbance on species richness as it was
found for impact survey of selective logging
\citep{Cannon1998, Baraloto2012}. Only the most disturbed plots had not
fully achieved the ongoing recovery of genus richness, although all had
already reached equivalent levels as those of some control plots. Such
long time richness recovery had already been observed in several logging
experiments, which also highlighted the role of random disturbance
damage in the richness decrease \citep{deAvila2015, Hu2018}.

>>>>

To counter the remaining taxonomic uncertainty plots taxonomic
composition and diversity were analysed at the genus level, \emph{i.e.}
referring to the genus of observed or trialed botanical names. The
analysis of taxonomic and functional composition of plots along time was
visualized in a two-dimensional ordination space based on non-metric
dimensional scaling of succesive floristic or functional inventories.
Plots trajectories along time was reported comparatively to the
inventories in 1989, 5 years after disturbance, which corresponded to
first inventory with a sufficient degree of uncertainty (\textless{}30\%
of undetermined trees). The inventories dissimilarity compared to the
reference 1989 inventory was reported using occurrence-based (Jaccard)
and abundance-based (Bray-Curtis) similarity measures. The trajectory of
inventories along time was visualized with the euclidean distance in the
two-dimensional ordination space to the 1989 inventory. The functional
trajectories of the leaf and stem and life traits were also visualized
with the community weighted means (CWM), representing the average trait
value in a community weighted by relative abundance of the species
carrying each value \citep{Diaz2007, Garnier2004}. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories.

@@

Over time, 828388 trees and 591 botanical species pertaining to 223
genus and 64 botanical families were recorded. Trajectories of taxonomic
and functional composition after disturbance were examined through the
ordination of successive inventories from 1989 (5 years after
disturbance) in the two-dimensional space from NMDS analysis based on
flora inventories and the corresponding functional composition.
Classifications were performed using either abundance-based Bray-Curtis
(Figure \ref{fig:NMDSplans}) or incidence-based Jaccard dissimilarity
(data not shown).

Both taxonomic and functional composition were substantially affected by
disturbance, and the maximum dissimilarity to 1984 reference inventory
composition was positively correlated with the disturbance intensity.
Until 30 years after disturbance, all disturbed plots taxonomic
composition remained significantly dissimilar to that of the 1989
reference inventory composition (Figure \ref{fig:NMDSplans}). All plots,
though, displayed a unimodal trajectory with a the return toward the
taxonomic and functional composition of the 1989 reference inventory
suggesting a shift towards a cyclic regime (Figure
\ref{fig:NMDSdivergence}). For taxonomic composition, the maximum
dissimilarity to 1984 inventory composition was reach around 20 years
after disturbance for the T2 and T3 plots and for one T1 plot while
control and other T1 plots continued to increase (Figure
\ref{fig:NMDSdivergence}.a). For functional composition, the maximum
dissimilarity to 1984 inventory composition was reach from 15 to 20
years after disturbance for the T2 and T3 plots while T1 and control
plots continued to increase (Figure \ref{fig:NMDSdivergence}.b). The
functional composition distancing from the reference inventory was also
positively correlated with disturbance intensity and similarly
stabilized or reduced (for 5 of T1, T2 and T3 plots) from 20 years after
disturbance (Figure \ref{fig:NMDSdivergence}).

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/NMDSplans-1}

@@

\caption{Trajectories of the plots in terms of \textbf{(a)} flora composition and \textbf{(b)} functional composition regarding the 6 leaf and stem functional traits,the maximum allometric height and seed mass class in the two-dimensional space from the NMDS performed for the 30 years after disturbance. Distance matrix for NMDS were computed from the Bray-curtis dissimilarity between successive inventories. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3).}\label{fig:NMDSplans}
\end{figure*}

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/NMDSdivergence-1}

@@

\caption{Trajectories of the distance to initial condition of the 30 sampled years in the two-dimensional space from the NMDS of v taxonomic composition at genus level and \textbf{(b)} functional composition. Distance are abundance-based Bary-Curtis metric. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3). The 0.025 and 0.975 percentile correspond to the variance observed for 50 iteration of the taxonomic uncertainty propagation and functional trait filling processes. }\label{fig:NMDSdivergence}
\end{figure}

@@

For all plots the trajectories of the community weighted means (CWM)
were drawn for the 8 functional and life history traits (Leaf thickness,
chlorophyll content, toughness and specific area,wood specific gravity
and barck thickness and seed mass and maximum adult height) (Figure
\ref{fig:CWM}).

@@

Except for leaf cholorphyll content, which displayed higher difference
between plots than among treatments which may be due to the completeness
of dataset, all CWM trajectories corresponded to significant changes
after disturbance. All functional traits and seed mass proportions
displayed a unimodal trajectories but with different times at maximum
and different values 30 years after disturbance. The weighted means of
communities specific maximum height at adult stage (\emph{Hmax}), leaf
toughness (\emph{L\_toughness}) and wood specific gravity (\emph{WD})
remained significantly lower than their initial value and than these of
the control plots (Figure \ref{fig:CWM}). The weighted means of bark
thickness (\emph{Bark\_thick}) similarly remained substantilly higher
than initially for all disturbed plots while the specfic leaf area
(\emph{SLA}) had almost recovered its initial value and this of the
undisturbed plots at the end of the experiment.

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/CWM-1}

@@

\caption{Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, \emph{L\_thickness}, chlorophyll content, \emph{L\_chloro}, toughness, \emph{L\_toughness} and specific area, \emph{SLA}), 2 stem traits (wood specific gravity, \emph{WD}, and bark thickness, \emph{Bark-thick}) and one life trait (Specific maximum height at adult stage, \emph{Hmax}). Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.}\label{fig:CWM}
\end{figure*}

@@

Trajectories of Richness, Shannon and Simpson taxonomic diversity were
examined at genus level in relation to the 1989 inventories (5 years
after disturbance) (Figure \ref{fig:DivTaxo}).

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/DivTaxo-1}

@@

\caption{Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities \textbf{(a)} Richness, \textbf{(b)} Shannon and \textbf{(c)} Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:DivTaxo}
\end{figure*}

While for undisturbed plots the Richness, Shannon and Simpson diversity
remained comparable to the values in 1989, the trajectories of all
disturbed plots showed significant increase positively correlated with
the disturbance intensity (\(\rho_{spearman}^{Shannon}=0.87\), and
\(\rho_{spearman}^{Simpson}=0.88\)) (Figure \ref{fig:DivTaxo}). The
Richness was less affected by the initial disturbance except for the
higher intensity (T3 plots) (Figure \ref{fig:DivTaxo}). The Shannon and
Simpson diversities however first significantly increased, illustrating
an increasing evenness, before reaching a maximum around 15 years after
disturbance for all plots except one from T1 treatment and one from T2
treatment (Figure \ref{fig:DivTaxo}).

@@

For all undisturbed plots the Rao diversity remained comparable to the
values in 1989, the trajectories of all disturbed plots showed
significant increase positively correlated with the disturbance
intensity (Figure \ref{fig:DivFun}). Although a first increase of plots
functional diversity that was all the more important than initial
disturbance was intense, 30 after disturbance the diversity was similar
among disturbed plots and significantly higher than this of control
plots.

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/DivFun-1}

@@

\caption{Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes.}\label{fig:DivFun}
\end{figure}

@@

The monitoring 30 years of disturbed forest communities confirmed the
limited direct impact of low disturbance on species richness as it was
found for impact survey of selective logging
\citep{Cannon1998, Baraloto2012a}. Only the most disturbed plots had not
fully achieved the ongoing recovery of genus richness, although all had
already reached equivalent levels as those of some control plots. Such
long time richness recovery had already been observed in several logging
experiments, which also highlighted the role of random disturbance
damage in the richness decrease \citep{deAvila2015, Hu2018}.
